---
title: "Transforming coordinates from crs to crs"
output: md_document
---

> history \| Create July 2023 \| Updated August 2023<br/>


<a href="https://polarwatch.noaa.gov/" target="_blank">NOAA PolarWatch</a> distributes gridded and tabular oceanographic data for polar regions.  Satellite data include geospatial information and most of them are in geographical coordinates (latitude and longitude).  Since the spatial coverage of many PolarWatch satellite data products, are within polar regions, the geospatial information is often represented in polar projected coordinates. (<a href="https://nsidc.org/data/user-resources/help-center/guide-nsidcs-polar-stereographic-projection" target="_blank">Polar Stereographic Projections</a>).  

When working with multiple satellite data sets from different crs coordinates, it makes it challenging to do data analysis and the coordinate transformation may be needed. 

In this exercise, we will learn to transform coordinates of one projection to another.

**The exercise demonstrates the following techniques:**
-   Downloading and saving a netcdf file from PolarWatch ERDDAP data server
-   Accessing satellite data and metadata in polar stereographic projection 
-   Convert the netcdf data into data.frame
-   Transforming coordinates using EPSG codes
-   Mapping data using the transformed coordinates


**Datasets **
    
   *Sea Ice Concentration*, NOAA/NSIDC Climate Data Record V4, Southern Hemisphere, 25km, Science Quality, 1978-2022, Monthly


**R Packages**

-   ncdf4 (reading data and metadata in netCDF format)
-   ggplot2, RColorBrewer, scales (mapping)
-   sf (projection)

### Install required packages

This code block will check if required packages are installed, and will install missing packages.

```{r setup_packages, message=FALSE, warning=FALSE}

# Function to check if pkgs are installed, install missing pkgs, and load
pkgTest <- function(x)
{
  if (!require(x,character.only = TRUE))
  {
    install.packages(x,dep=TRUE,repos='http://cran.us.r-project.org')
    if(!require(x,character.only = TRUE)) stop(x, " :Package not found")
  }
}

list.of.packages <- c("ncdf4" , "sf", "ggplot2","scales", "RColorBrewer")

# create list of installed packages
pkges = installed.packages()[,"Package"]
for (pk in list.of.packages) {
  pkgTest(pk)
}

```

## Antarctic Sea Ice Concentration Data

The sea ice concentration data are used to calculate the spatical coverage of sea ice.  The data product in this exercise is sea ice concentration data of the southern hemisphere, and is produced by the NOAA/NSIDC from passive microwave sensors using the Climate Data Record (CDR).  The data are projected in the southern polarstereographic projection (**EPSG:3031**). 

This sea ice concentration data values in the dataset are represented as proportions ranging from 0 to 1. The resolution is 25km, meaning each grid in this data set represents a value that covers a 25km by 25km area.  Please note that other values beyond 1 represents fill values.  More detailed information can be found in the NSIDC Data Product User Guide (linke below)

**Userful links**

-   <a href="https://polarwatch.noaa.gov/catalog/ice-sq-nh-nsidc-cdr-v4/preview/?dataset=daily&var=cdr_seaice_conc&time_min=2022-05-31T00:00:00Z&time_max=2022-05-31T00:00:00Z&proj=epsg3413&colorBar=KT_ice|||0|1|" target="_blank">NOAA PolarWatch Data Product Page (download, preview)</a>
-   <a href="https://nsidc.org/data/g02202/versions/4" target="_blank">NSIDC Data Product Description</a>
-   <a href="https://nsidc.org/sites/default/files/g02202-v004-userguide_1_1.pdf" target="_blank">NSIDC Data Product User Guide (pdf)</a>


### Get data from ERDDAP 
```{r}
# download the sea ice data NetCDF file
url <- "https://polarwatch.noaa.gov/erddap/griddap/nsidcG02202v4shmday.nc?cdr_seaice_conc_monthly[(2022-12-01T00:00:00Z):1:(2022-12-01T00:00:00Z)][(4350000.0):1:(-3950000.0)][(-3950000.0):1:(3950000.0)]"

sic <- download.file(url, destfile="sic.nc", mode='wb')

# file open
ds <- nc_open('sic.nc')

# print metadata
#print(ds)

# get data into r variables 
xgrid <- ncvar_get(ds, "xgrid")
ygrid <- ncvar_get(ds, "ygrid")
sic <- ncvar_get(ds, "cdr_seaice_conc_monthly")  #lat and lon
dim(sic)


# close 
nc_close(ds)
```
### Map polar stereographic projected data 
```{r}

# create dataframe
sicd <- expand.grid(xgrid=xgrid, ygrid=ygrid)
sicd$sic <- array(sic, dim(xgrid)*dim(ygrid))

# exclude fillvalue
sicd$sic[sicd$sic > 2] <- NA 

# map sea ice concentration
ggplot(data = sicd, aes(x = xgrid, y = ygrid, fill=sic) ) + 
       geom_tile() + 
       coord_fixed(ratio = 1) + 
       scale_y_continuous(labels=comma) + 
       scale_x_continuous(labels=comma) +
       scale_fill_gradientn(colours=rev(brewer.pal(n = 3, name = "Blues")),na.value="tan")+       ggtitle("Sea Ice Concentration on Polar Steregraphic projection")
```


## Transforming from CRS to CRS

When transforming from one CRS to another, it is important to inspect CRS definitions and the transformation function for proper transformation.
We will transform from CRS EPSG: 3031 (NSIDC Polar Stereographic South) to EPSG: 4326 (geographic coordinate system) 

```{r}

xy_sf <- st_as_sf(sicd, coords=c("xgrid", "ygrid"), crs=3031)

# st_transform output order: lat , lon
latlon_sf <- st_transform(xy_sf, crs=4326)

latlon_df <- as.data.frame(st_coordinates(latlon_sf))
names(latlon_df) <- c("Lat", "Lon")
seaiceconc <- sicd$sic
# create dataframe and add names
sicdf_latlon <- data.frame(cbind(latlon_df,seaiceconc))
head(na.omit(sicdf_latlon), 5)


```
### Plot data with new coordinates on a gloal map

```{r}
sic_sf <- st_as_sf(sicdf_latlon, coords = c('Lat', 'Lon'))
st_crs(sic_sf) = "4326"

# plot sea ice concentration on an unprojected map
ggplot(sic_sf) + geom_sf(aes(color = seaiceconc)) + labs(title = "Sea Ice Concentration of Antarctica")



```


