---
title: "subset-polar-data-with-shapefile"
author: "Sunny Hospital"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### The tutorial demonstrates the following techniques
* Download sea ice satellite data from the PolarWatch ERDDAP data server
* Import geographical features of Lake Iliamna from a shapefile
* Transform data from one projection to another
* Subset the satellite data for Lake Iliamna
* Visualize data in different projection

```{r read_shapefile}
# Set directory path for shapefile
dir_path <- '../resources/'

# Import shape files (Longhurst coordinates)
shapes <- read_sf(dsn = dir_path, layer = "Longhurst_world_v4_2010")

# Example List of all the province names
shapes$ProvCode

# Get boundary coordinates for Gulf Stream region (GFST)
GFST <- shapes[shapes$ProvCode == "GFST",]

xcoord <- st_coordinates(GFST)[,1]
ycoord <- st_coordinates(GFST)[,2]
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
