---
title: Virtual Buoy Example 
author: NOAA CoastWatch
date: August 2023
output:
  md_document
---

<<<<<<< HEAD
=======
# Virtual Buoy example

>>>>>>> sun-dev
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.path = "images/"
)
```

> Updated August 2023 <br/>

<<<<<<< HEAD
There are buoys in many locations around the world that provide data streams of oceanic and atmospheric parameters. The data are often available through data centers like the those National Data Buoy Center (NDBC https://www.ndbc.noaa.gov), ARGO floats program (http://www.argo.ucsd.edu) or ERDDAP data server (https://www.our server). In situ buoy data are widely used to monitor environmental conditions.  In the absence of in situ buoy data, whether the buoy operation is discontinued, interrupted, or limited, satellite data with within temporal and spatial coverage can be used to create a dataset in a buoy data format.  

To demonstrate the satellite data being used in place of buoy data, we will use sea surface temperature data which both satellite and buoy collect.  

## Objective

This tutorial will demonstration how to transform satellite data into a buoy data format

## The tutorial demonstrates the following techniques

* Download the satellite data and bouy data of the time period and spatial coverage
* Visualize the datasets
* Transform the satellite data into a buoy data format
* Validate the transformed satellite data with the actual buoy data
* Plot the transformed satellite data and the buoy data
* Perform a linear regression of satellite vs. buoy data
* Plot satellite vs. buoy data and overlay the regression line

## Datasets used

-  <a href="https://coastwatch.pfeg.noaa.gov/erddap/griddap/nesdisBLENDEDsstDNDaily.graph">The sea surface temperature (SST) satellite data</a> from NOAA Geo-polar blended analysis are used for transforming to buoy data format
  * Sea-Surface Temperature, NOAA Geo-polar Blended Analysis Day+Night, GHRSST,Near Real-Time, Global 5km, 2019-Present, Daily*  

-  <a href="https://coastwatch.pfeg.noaa.gov/erddap/tabledap/cwwcNDBCMet.graph?time%2Cwtmp%2Cwd&station=%2246259%22&time%3E=2020-09-15T00%3A00%3A00Z&time%3C=2022-09-15T00%3A00%3A00Z&.draw=markers&.marker=5%7C5&.color=0x000000&.colorBar=%7C%7C%7C%7C%7C&.bgColor=0xffccccff"> NDBC Standard Meteorological Buoy Data </a> will be used to compare with the satellite data.
  * Buoy Station No. **46259** 
  * Location: Off the California coast at 34.737N latitude and 121.664E longitude 
=======
There are buoys in many locations around the world that provide data streams of oceanic and atmospheric parameters. The data are often available through data centers like the those National Data Buoy Center (NDBC https://www.ndbc.noaa.gov), ARGO floats program (http://www.argo.ucsd.edu) or ERDDAP data server (https://coastwatch.pfeg.noaa.gov/erddap/). In situ buoy data are widely used to monitor environmental conditions.  In the absence of in situ buoy data, whether the buoy operation is discontinued, interrupted, or limited, satellite data with within temporal and spatial coverage can be used to create a dataset in a buoy data format.  


## Objective

This tutorial will demonstration how to shape satellite data into a buoy data format

## The tutorial demonstrates the following techniques

* Downloading the satellite and buoy data from ERDDAP data server
* Visualizing the datasets
* Reshaping the satellite data into a buoy data format
* Resampling buoy data (aggregation) to match satellite data temporal resolution
* Validating the  satellite data with the actual buoy data
* Performing a linear regression of satellite vs. buoy data
* Creating residual plot 

## Datasets used

  **Sea-Surface Temperature, NOAA Geo-polar Blended Analysis Day+Night, GHRSST,Near Real-Time, Global 5km, 2019-Present, Daily**  
<a href="https://coastwatch.pfeg.noaa.gov/erddap/griddap/nesdisBLENDEDsstDNDaily.graph" target="_blank">The sea surface temperature (SST) </a> satellite data from NOAA Geo-polar blended analysis 

**NDBC Standard Meteorological Buoy Data, 1970-present**
<a href="https://coastwatch.pfeg.noaa.gov/erddap/tabledap/cwwcNDBCMet.graph?time%2Cwtmp%2Cwd&station=%2246259%22&time%3E=2020-09-15T00%3A00%3A00Z&time%3C=2022-09-15T00%3A00%3A00Z&.draw=markers&.marker=5%7C5&.color=0x000000&.colorBar=%7C%7C%7C%7C%7C&.bgColor=0xffccccff"> NDBC Standard Meteorological Buoy Data </a> from the buoy station no. **46259** are from off the California coast at 34.737N latitude and 121.664E longitude. 
 
>>>>>>> sun-dev

## References

- <a href="https://coastwatch.pfeg.noaa.gov/data.html">NOAA CoastWatch Westcoast Node Data Catalog</a>
- <a href="https://www.ndbc.noaa.gov/download_data.php?filename=46259h2017.txt.gz&dir=data/historical/stdme ">NOAA National Data Buoy Center</a>

## Install required packages and load libraries
```{r install,message=FALSE,warning=FALSE}

# Function to check if pkgs are installed, and install any missing pkgs
pkgTest <- function(x)
{
  if (!require(x,character.only = TRUE))
  {
    install.packages(x,dep=TRUE,repos='http://cran.us.r-project.org')
    if(!require(x,character.only = TRUE)) stop(x, " :Package not found")
  }
}


# Create list of required packages
list.of.packages <- c("utils", "ggplot2", "dplyr", "lubridate")

# Create list of installed packages
pkges = installed.packages()[,"Package"]

# Install and load all required pkgs
for (pk in list.of.packages) {
  pkgTest(pk)
}

```


<<<<<<< HEAD
## Download NDBC buoy data (August 16 2020 - August 16 2022)
The virtual buoy data with Station ID #46259 will be downloaded from ERDDAP as the ground truth.  

```{r}

=======
## Download NDBC buoy data

We will download NDBC buoy data between August 16, 2020 and August 16, 2022 from the CoastWatch ERDDAP server. 

The data can be downloaded by sending a data request to the ERDDAP server via URL.  The data request URL includes a dataset ID of our interest and other query conditions if subset of the data product is of interest. 

To learn more about how to set up ERDDAP URL data request, please go to <a href="" target="_blank">ERDDAP module page</a>.


```{r}

# Set ERDDAP data request URL
>>>>>>> sun-dev
buoy_url <- "https://coastwatch.pfeg.noaa.gov/erddap/tabledap/cwwcNDBCMet.csv?time%2Clongitude%2Clatitude%2Cwtmp&station%3E=%2246259%22&station%3C=%2246259%22&time%3E=2020-08-16T00%3A00%3A00Z&time%3C=2022-08-16T17%3A52%3A00Z"

# Set file name
fname = 'buoy.csv'

# Download file
download.file(buoy_url, fname)

<<<<<<< HEAD
# Read into data frame, skip the first row
=======
# Read into data frame, skip first 2 rows that contain variable names and units
>>>>>>> sun-dev
buoy_df <- read.csv(fname, skip=2, header=TRUE)

# Name the columns
names(buoy_df) <- c("utc", "lon", "lat", "sst" )

<<<<<<< HEAD
# Add formatted data column
=======
# Add additional date column
>>>>>>> sun-dev
buoy_df$date <- as.Date(buoy_df$utc, tz = "UTC")

# Show the first 3 rows
head(buoy_df, 3)
<<<<<<< HEAD

```
## Visualize SST from the Buoy 
```{r}
options(repr.plot.width = 10)

ggplot(buoy_df, aes(x = date, y = sst)) +
  geom_line(color='blue') +
 # geom_point(size=.2, color='red')+
  theme(axis.text.x = element_text(angle = 90),plot.title=element_text(hjust=0.5))+
   labs(x="Date", y="Sea Surface Temp (Celcius)", title="SST from NDBC Buoy Station: 46259 (Aug 2022- Aug 2023) ")
  
=======
```
## Visualize SST from the Buoy 

The plot shows the downloaded sea surface temperature by date.  
```{r}
# Plot sea surface temperature values from buoy
p <- ggplot(buoy_df, aes(x = date, y = sst)) +
    geom_line(color='blue') +
    theme(axis.text.x = element_text(angle = 90),plot.title=element_text(hjust=0.5))+
    labs(x="Date", y="Sea Surface Temp (Celcius)", title="Sea surface temperature from Buoy")
p
>>>>>>> sun-dev
```

## Download the Satellite Sea Surface Temperature (SST) Data 

the Sea Surface Temperature (SST) is the NOAA GeoPolar Blended SST dataset (in Celcius) from many satellite sensors to obtain good daily coverage of the globe at 5km resolution, and then an interpolation method is applied to fill in data gaps. 

The data request can be sent to CoastWatch ERDDAP server via URL with a query string to specify the temporal and spatial coverage of our interest.  In this case, we want to subset the satellite data to match the buoy station location.

To learn more about ERDDAP data request via URL, please go to <a href="">ERDDAP module</a>.

<<<<<<< HEAD
### Sending data request to CoastWatch ERDDAP Server
```{r}
# Set ERDDAP URL with subset query string
=======
### Send satellite data request to CoastWatch ERDDAP Server
```{r}
# Set ERDDAP URL for the satellite data
>>>>>>> sun-dev
url <- "https://coastwatch.pfeg.noaa.gov/erddap/griddap/nesdisBLENDEDsstDNDaily.csv?analysed_sst%5B(2020-08-16T12:00:00Z):1:(2022-08-16T12:00:00Z)%5D%5B(34.737):1:(34.737)%5D%5B(-121.664):1:(-121.664)%5D"

# Set file name
fname = 'sst.csv'

# Download file
download.file(url, fname)

<<<<<<< HEAD
# Read into data frame, skip first 2 rows that contain 
sst_df <- read.csv(fname, skip=2, header=TRUE)
names(sst_df) <- c("utc", "lat", "lon", "sst")
# Add formatted data column
sst_df$date <- as.Date(sst_df$utc, tz = "UTC")
=======
# Read into data frame, skip first 2 rows that contain variable names and units
sst_df <- read.csv(fname, skip=2, header=TRUE)
names(sst_df) <- c("utc", "lat", "lon", "sst")

# Add formatted data column
sst_df$date <- as.Date(sst_df$utc, tz = "UTC")

>>>>>>> sun-dev
# Show the first 3 rows
head(sst_df, 3)
```
## Clean up the data
Apply a conservative allowable data range.
For the lower end of the range, the freezing point of seawater (ca. -2).
For the high end of the range, value unlikely to be seen in the area of interest (e.g. 45 degrees C).

```{r}
<<<<<<< HEAD
# Remove outliers sst values outside of -2 and 45 deg C
=======
# Filter data that are within the valid range (-2 and 45)
>>>>>>> sun-dev
sst_df_clean = sst_df %>%
  filter(sst >=-2 & sst <= 45)

```

## Visualize SST data from the Satellite Data
```{r}
<<<<<<< HEAD

ggplot(sst_df, aes(x = date, y = sst)) +
  geom_line(color='blue') +
  geom_point(size=.5, color='red')+
  theme(axis.text.x = element_text(angle = 90),plot.title=element_text(hjust=0.5))+
   labs(x="Date", y="Sea Surface Temp (Celcius)", title="2022 Aug- 2023 Aug Sea Surface Temperature at 34.736 deg North and -121.664 deg East")
  
=======
# Plot sea surface temperature values from satellite
p <- ggplot(sst_df, aes(x = date, y = sst)) +
  geom_line(color='blue') +
  geom_point(size=.5, color='red')+
  theme(axis.text.x = element_text(angle = 90),plot.title=element_text(hjust=0.5))+
   labs(x="Date", y="Sea Surface Temp (Celcius)", title="Sea surface temperature from satellite")
p
>>>>>>> sun-dev
```


## Resample the buoy data to match satellite data

The sampling resolution for the buoy data is a sample every 30 minutes. However, the temporal resolution for the satellite dataset is daily. We will downsample the buoy data by computing daily mean to match the temporal resolution of the satellite data.

```{r}
<<<<<<< HEAD
=======
# Aggregating (taking a mean value) grouped by day
>>>>>>> sun-dev
buoy_ds <- buoy_df %>%
  group_by(date = floor_date(date, unit="days")) %>%
  summarise(mean_sst = mean(sst))

<<<<<<< HEAD
=======
# Show first 3 rows
>>>>>>> sun-dev
head(buoy_ds, 3)
```


## Clean up the data
Apply a conservative allowable data range.
For the lower end of the range, the freezing point of seawater (ca. -2).
For the high end of the range, value unlikely to be seen in the area of interest (e.g. 45 degrees C).

```{r}
# Remove outliers sst values outside of -2 and 45 deg C
buoy_ds_clean = buoy_ds %>%
  filter(mean_sst >=-2 & mean_sst <= 45)
<<<<<<< HEAD

=======
>>>>>>> sun-dev
```


## Visualize downsampled buoy data
```{r}

<<<<<<< HEAD
=======
# Plot daily mean sst
>>>>>>> sun-dev
ggplot(buoy_ds_clean, aes(x = date, y = mean_sst)) +
  geom_line(color='blue') +
  geom_point(size=.5, color='red')+
  theme(axis.text.x = element_text(angle = 90),plot.title=element_text(hjust=0.5))+
   labs(x="Date", y="Sea Surface Temp (Celcius)", title="2022 Aug- 2023 Aug Downsampled Buoy SST")
  
```

## Merge Satellite and Buoy data
We will use dplyr::inner_join() function to merge two data frame (satellite and buoy data) based on the dates appear on both dataframes.


```{r}
<<<<<<< HEAD
=======
# Combine two data frames with date column where dates exist in both data frame
>>>>>>> sun-dev
merged_df <- inner_join(sst_df_clean[c("date", "sst")],  buoy_ds_clean, by = "date")
```

## Plot both satellite and buoy data

```{r}
<<<<<<< HEAD

=======
# Plot satellite sst and buoy daily mean sst 
>>>>>>> sun-dev
p <- ggplot(merged_df, aes(x = date)) +  
    geom_line(aes(y = sst, color = "Satellite")) +   
    geom_line(aes(y = mean_sst, color = "NDBC Buoy")) +   
    scale_color_manual(name = "Data source", 
                       values = c("Satellite" = "blue", "NDBC Buoy" = "orange"))+
    labs( x = "Date", y = "Temperature (in Celcius)", title="SST from Satellite and NDBC Buoy" )+
    theme(axis.text.x = element_text(angle = 0),plot.title=element_text(hjust=0.5))
<<<<<<< HEAD
 
=======
>>>>>>> sun-dev
p
```
## Perform a simple linear regression 

```{r}
<<<<<<< HEAD
model <- lm(mean_sst ~ sst, data = merged_df)   
```

## Plot fiited vs. residuals

```{r}

=======

# Run linear regression 
model <- lm(mean_sst ~ sst, data = merged_df)   
```

## Plot fitted vs. residuals

```{r}
# Plot fitted vs residuals 
>>>>>>> sun-dev
ggplot() +
  geom_point(aes(x = model$fitted.values, y = model$residuals), size=0.8) +
  geom_hline(yintercept = 0, color = "red") +
  labs(x = "Fitted Values", y = "Residuals", title = "Residual Plot") +
  theme(plot.title=element_text(hjust=0.5))
```