---
title: "Mask shallow water pixels for satellite ocean color datasets"
author: "hui.shi@noaa.gov"
date: "2023-11-21"
output: 
  md_document:
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Tutorial: Mask shallow water pixels for satellite ocean color datasets

Satellite sensed ocean color data in the coastal shallow water can be contaminated by bottom reflectance. Here we use a high resolution bathymetry data to generate a mask to remove ocean color pixels that contain a portion (more than 5%) of shallow water area (less than 30m) (Couch et al., 2023).

The bathymetry data is the ETOPO Global Relief Model integrates topography, bathymetry, and shoreline data version 2022 in 15 arc-second resolution  (https://www.ncei.noaa.gov/products/etopo-global-relief-model).

The ocean color data is the ESA CCI monthly chlorophyll-a concentration (https://climate.esa.int/en/projects/ocean-colour).



Reference: Couch CS, Oliver TA, Dettloff K, Huntington B, Tanaka KR and Vargas-Ángel B (2023) Ecological and environmental predictors of juvenile coral density across the central and western Pacific. Front. Mar. Sci. 10:1192102. doi: 10.3389/fmars.2023.1192102

```{r}
### Load libraries
library(rerddap)
library(raster) 
library(sp) 
library(cmocean)
library(here)
library(ncdf4)
```

```{r}
# This is where the data are and where the plots will go
Dir <- here()
```

```{r}

# Thanks to Kisei Tanaka, the tutorial is modified based on the scripts he developed for the EDS. And part of the tutorial is modified based on the previous version developed by Melanie Abecassis and Thomas Oliver.

### Load data

# Bounding box for Oahu:
lon_range = c(-158.39+360, -157.55+360)
lat_range = c(21.14, 21.8)

# Access bathymetry and chl-a data from OceanWatch ERDDAP server
ERDDAP_Node = "https://oceanwatch.pifsc.noaa.gov/erddap/"

ETOPO_id = 'ETOPO_2022_v1_15s'
ETOPO_info=info(datasetid = ETOPO_id,url = ERDDAP_Node)
bathy =  griddap(url = ERDDAP_Node, ETOPO_id, 
                 latitude = lat_range, longitude = lon_range)

# Download ocean color to local as a netCDF file
CCI_id = 'esa-cci-chla-monthly-v6-0'
CCI_info=info(datasetid = CCI_id,url = ERDDAP_Node)
var=CCI_info$variable$variable_name
chl = griddap(url = ERDDAP_Node, CCI_id, 
                   time = c('1998-01-01', '2022-12-01'),
                   latitude = lat_range, longitude = lon_range,
                   fields = var[1],
                   store=disk('chl_data'))
```

```{r}
# Take a look at bathymetry data
r_bathy=raster(bathy$summary$filename)

plot(r_bathy,main="ETOPO Bathymetry (m)")
contour(r_bathy,levels=c(-30,-1000,-2000),add=TRUE) 
```

```{r}
# Take a look at chlorophyll-a data
r_chl=raster(chl$summary$filename,varname=var[1]) 

plot(log(r_chl),main="ESA CCI Chl-a (log scale)",col=cmocean('algae')(50))
contour(r_bathy,levels=c(-30,-1000,-2000),add=TRUE) 
```

```{r}
### Create a mask for shallow water pixels
#Convert raster bathymetry to dataframe for counting
df_bathy = data.frame(rasterToPoints(r_bathy))
coordinates(df_bathy) <- ~x+y
crs(df_bathy) = crs(r_chl[[1]])

# Examine number of shallow water pixels in each chl-a grid point
count_shallow_pixels=function(depths,threshold=-30,na.rm=F){ 
  return(length(which(depths>threshold))) 
} 
N_shallow = rasterize(x = df_bathy,y=r_chl,fun=count_shallow_pixels)[[2]]
plot(N_shallow,main="N Shallow Pixels", col=cmocean('balance')(50))
```

```{r}
# Examine percentage of shallow water pixels in each chl-a grid point
percent_shallow_pixels=function(depths,threshold=-30, na.rm=F){ 
  return(length(which(depths>threshold))/length(depths)) 
} 
per_shallow =  rasterize(x = df_bathy,y=r_chl,fun=percent_shallow_pixels)[[2]]
plot(per_shallow,main="% Shallow Pixels", col=cmocean('amp')(50))
```

```{r}
# Set a percentage threshold to create the shallow water pixel mask
percent_threshold = 0.05
depth_mask = r_chl/r_chl
depth_mask[,]= 1
depth_mask[per_shallow>= percent_threshold]= NA
plot(depth_mask,main="Shallow water mask")
```

```{r}
### Apply created shallow water mask to chl-a data and compare unmasked and masked maps 
# Read in the files previousely downloaded
files = list.files('chl_data/', full.names = T)
# Read the file into R and make it to rasterstack
stack_chl = stack(files)
# Convert raster data to dataframe for calculating climatology
df_chl = as.data.frame(rasterToPoints(stack_chl))
df_chl$z = rowMeans(df_chl[,3:dim(df_chl)[2]], na.rm = T)
# Convert dataframe to raster for mapping
r_chl_clim = rasterFromXYZ(df_chl[,c("x", "y", "z")])
# Map unmasked climatology
plot(log(r_chl_clim),main="Chl-a (log scale)",col=cmocean('algae')(50))
contour(r_bathy,levels=c(-30,-1000,-2000),add=TRUE) 
```

```{r}
# Apply Mask, calculate climatology and map
r_chl_masked = mask(x = stack_chl, mask = depth_mask)
# Convert masked raster data to dataframe for calculating climatology
df_chl_masked = as.data.frame(rasterToPoints(r_chl_masked))
df_chl_masked$z = rowMeans(df_chl_masked[,3:dim(df_chl_masked)[2]], na.rm = T)
# Convert dataframe to raster for mapping
r_chl_masked_clim = rasterFromXYZ(df_chl_masked[,c("x", "y", "z")])
# Map masked climatology
plot(log(r_chl_masked_clim),main="Masked Chl-a (log scale)",col=cmocean('algae')(50))
contour(r_bathy,levels=c(-30,-1000,-2000),add=TRUE) 

```

```{r}
###Output masked chl-a to netcdf file
# Grab var name and unit from unmasked nc file
nc = nc_open(paste0(files))
variable_name = as.character(nc$var[[1]][2])
variable_unit = as.character(nc$var[[1]][8])
x_name = nc$dim$longitude$name
y_name = nc$dim$latitude$name
z_name = nc$dim$time$name
z_unit = nc$dim$time$units
nc_close(nc)
# Set a file name and output path for the masked data
masked_fln = 'esa_cci_monthly_chl-a_masked.nc'
mask_path = paste0(Dir,  "/output/")
if (!dir.exists(mask_path)) {
  dir.create(mask_path)
}
# Write out masked nc.file
writeRaster(r_chl_masked,
            paste0(mask_path, masked_fln),
            overwrite = T,
            varname = variable_name,
            varunit = variable_unit,
            xname = x_name,
            yname = y_name,
            zname = z_name,
            zunit = z_unit)
```

