---
title: Match up satellite to buoy data
author: NOAA CoastWatch
date: March 2024
output:
  md_document:
    variant: gfm
---
```{r echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.path = "images/",
  warning = FALSE, message = FALSE
)
```  
  
# Match up satellite and buoy data
> history | Updated March 2024

## Background
There are buoys in many locations around the world that provide data streams of oceanic and atmospheric parameters. The data are often available through data centers like the National Data Buoy Center (NDBC https://www.ndbc.noaa.gov) and the ARGO floats program (http://www.argo.ucsd.edu). In situ buoy data are widely used to monitor environmental conditions.

In-situ buoy data can be used to evaluate the accuracy of satellite data.

## Objective
In this exercise, we will learn how to match up satellite data to in situ buoy data using rerddap and rxtracto R packages.  


## The exercise demonstrates the following techniques:
  
* Downloading tabular data (buoy data) from CoastWatch ERDDAP data server
* Retrieving information about a dataset from ERDDAP
* Subsetting satellite data within a rectangular boundary
* Matching satellite data with the buoy data
* Running statistical analysis to compare buoy and satellite data
* Producing satellite maps and overlaying buoy data

## Datasets used:
  
-  <a href="https://coastwatch.pfeg.noaa.gov/erddap/griddap/nesdisBLENDEDsstDNDaily.graph">The sea surface temperature (SST) satellite data</a> from the NOAA Geo-polar blended analysis
-  <a href="https://coastwatch.pfeg.noaa.gov/erddap/tabledap/cwwcNDBCMet.graph?time%2Cwtmp%2Cwd&station=%2246259%22&time%3E=2020-09-15T00%3A00%3A00Z&time%3C=2022-09-15T00%3A00%3A00Z&.draw=markers&.marker=5%7C5&.color=0x000000&.colorBar=%7C%7C%7C%7C%7C&.bgColor=0xffccccff"> The NDBC Standard Meteorological Buoy Data (dataset ID: cwwcNDBCMet) </a> is used for validating or ground truthing the satellite SST data


  
## Install required packages and load libraries  
  
```{r install, message=FALSE, warning=FALSE}

# Function to check if pkgs are installed, install missing pkgs, and load

pkgTest <- function(x)
{
  if (!require(x,character.only = TRUE))
  {
    install.packages(x,dep=TRUE)
    if(!require(x,character.only = TRUE)) stop("Package not found")
  }
}

list.of.packages <- c( "ncdf4", "rerddap", "plotdap", "httr",
                       "lubridate", "gridGraphics",  "mapdata",
                       "ggplot2", "RColorBrewer", "grid", "PBSmapping", 
                       "rerddapXtracto","dplyr","viridis","cmocean")

# create list of installed packages
pkges = installed.packages()[,"Package"]

for (pk in list.of.packages) {
  pkgTest(pk)
}

```

## Downloading buoy data from ERDDAP  
**Extract data using the rerddap::tabledap function**  
  
  Using rerddap::tabledap function, we will request and download data with the following specifications:
  
* Buoy dataset ID: cwwcNDBCMet
* Region boundaries: 35 to 40 north latitude and -125 to -120 east longitude  
* Time span: 08/01/2023 to 08/10/2023
* Variables:  station, latitude, longitude, time, and water temperature parameters  

```{r tabledap}
# Subset and download tabular data from ERDDAP 

ERDDAP_Node = "https://coastwatch.pfeg.noaa.gov/erddap"

NDBC_id = 'cwwcNDBCMet'
NDBC_info=info(datasetid = NDBC_id,url = ERDDAP_Node)

buoy <- rerddap::tabledap( url = ERDDAP_Node, NDBC_id,
  fields=c('station', 'latitude',  'longitude', 'time', 'wtmp'), 
  'time>=2023-08-01',   'time<=2023-08-10', 
  'latitude>=35','latitude<=40', 'longitude>=-125','longitude<=-120',
  'wtmp>0'
)

#Create data frame with the downloaded data
buoy.df <-data.frame(station=buoy$station,
                     longitude=as.numeric(buoy$longitude),
                     latitude=as.numeric(buoy$latitude),
                     time = buoy$time,
                     temp=as.numeric(buoy$wtmp))

# Check for unique stations
unique.sta <- unique(buoy$sta)
n.sta <- length(unique.sta)
n.sta
 
```


**Plot the buoy data for the first 10 stations in buoy.df**
  
  Let's see what the buoy data looks like for our time period.

```{r buoy_plot}

plot(buoy.df$time,buoy.df$temp,type='n', xlab='Date', ylab='SST (ºC)',main='SST from the first 10 stations')


for (i in 1:10){
  I=which(buoy.df$station==unique.sta[i])
  lines(buoy.df$time[I],buoy.df$temp[I])
}
```

**Select buoy data closest in time to satellite data**  

Since buoy data are hourly and the satellite data are daily, the buoy data needs to be averaged daily for each station.

```{r average}

buoy.df.day=buoy.df %>%
  mutate(date = floor_date(time, unit="days")) %>%
  group_by(station, date) %>%
  summarize(
    lon=mean(longitude),
    lat=mean(latitude),
    temp.day=mean(temp),
    .groups="drop"
    )

head(buoy.df.day)
```

## Download Satellite SST (sea surface temperature) data

We will use Sea surface temperature (SST) satellite data from CoastWatch West code node ERDDAP server.

URL: https://coastwatch.pfeg.noaa.gov/erddap/
Dataset ID:nesdisBLENDEDsstDNDaily 

```{r dataInfo}

url=  'https://coastwatch.pfeg.noaa.gov/erddap/'
datasetid = 'nesdisBLENDEDsstDNDaily'

# Get Data Information given dataset ID and URL
dataInfo <- rerddap::info(datasetid, url)

# Show data Info
dataInfo

```

**Extract the matchup data using rxtracto**  

We will extract satellite data for each buoy station.  
1. get coordinates of each buoy station
2. use the rxtracto function and the buoy coordinates to download satellite data closest to each station

```{r}
# Set the variable name of interest from the satellite data
parameter <- 'analysed_sst'

# Set x,y,t,z coordinates based on buoy data
xcoord <- buoy.df.day$lon
ycoord <- buoy.df.day$lat
tcoord <- buoy.df.day$date
```

```{r matchups}

# Set the variable name of interest from the satellite data
parameter <- 'analysed_sst'

# Set x,y,t,z coordinates based on buoy data
xcoord <- buoy.df.day$lon
ycoord <- buoy.df.day$lat
tcoord <- buoy.df.day$date

# Extract satellite data 
extract <- rxtracto(dataInfo, parameter=parameter, 
                    tcoord=tcoord,
                    xcoord=xcoord,
                    ycoord=ycoord,
                    xlen=.01,ylen=.01)
                     
buoy.df.day$sst<-extract$`mean analysed_sst`

```

**Get subset of data where a satellite value was found**  

* Our satellite product is gap-free (gaps due to clouds were filled using some interpolation) but it has a spatial resolution of 5km. Some buoy stations may be so close to shore that they end up in the landmask of the satellite data. So let's find the stations that were matched up to an SST pixel. 

```{r goodvalues}
# Get subset of data where there is a satellite value 
goodbuoy<-subset(buoy.df.day, sst > 0)
unique.sta<-unique(goodbuoy$station)
nbuoy<-length(unique.sta)
ndata<-length(goodbuoy$station)
```

## Compare results for satellite and buoy
* Plot the satellite SST verses the buoy temperature to visualize how well the two datasets match each other.

```{r xyplot}
# Set up map title 
main="California coast, 8/1/23-8/10/23"   

p <- ggplot(goodbuoy, aes(temp.day, sst,color=lat)) + 
     coord_fixed(xlim=c(8,25),ylim=c(8,25)) 
p + geom_point() + 
  ylab('Satellite SST')  + 
  xlab('Buoy average daily SST') +
  scale_x_continuous(minor_breaks = seq(8, 25)) + 
  scale_y_continuous(minor_breaks = seq(8, 25)) + 
  #geom_abline(a=fit[1],b=fit[2]) +
  #annotation_custom(my_grob) + 
  #scale_color_gradientn(colours = "viridis", name="Buoy\nLatitude") +
  scale_color_viridis(discrete = FALSE, name="Buoy\nLatitude") +
  labs(title=main) + theme(plot.title = element_text(size=20, face="bold", vjust=2)) 

```

Run a linear regression of Blended SST versus the buoy data. 
* The R squared is close to 1 (0.8733)
* The slope is 0.7151
```{r lg}
lmHeight = lm(sst~temp.day, data = goodbuoy)
summary(lmHeight)
```
## Create a map of SST and overlay the buoy data
**Extract blended SST data for August 1st, 2023**


```{r getviirs}
# First define the box and time limits of the requested data 
ylim<-c(32,42)
xlim<-c(-127,-118)

# Extract the monthly satellite data
SST <- rxtracto_3D(dataInfo,xcoord=xlim,ycoord=ylim,parameter=parameter, 
                   tcoord=c('2023-08-01','2023-08-01'))

SST$sst <- drop(SST$analysed_sst)
```

**Create the map frame for the satellite data and buoy SST overlay**  

```{r mapsetup}
mapFrame<- function(longitude,latitude,sst){
  dims<-dim(sst)
  sst<-array(sst,dims[1]*dims[2])
  sstFrame<-expand.grid(x=longitude,y=latitude)
  sstFrame$sst<-sst
  return(sstFrame)
}

sstFrame<-mapFrame(SST$longitude,SST$latitude,SST$sst)
coast <- map_data("worldHires", ylim = ylim, xlim = xlim)
my.col <- colorRampPalette(rev(brewer.pal(11, "RdYlBu")))(22-13) 

buoy2<-subset(buoy.df.day, month(date)==8 &day(date)==1 & temp.day > 0)
```

**Create the map**   

```{r map}

myplot<-ggplot(data = sstFrame, aes(x = x, y = y, fill = sst)) +
  geom_tile(na.rm=T) +
  geom_polygon(data = coast, aes(x=long, y = lat, group = group), fill = "grey80") +
  theme_bw(base_size = 15) + ylab("Latitude") + xlab("Longitude") +
  coord_fixed(1.3,xlim = xlim, ylim = ylim) +
  scale_fill_cmocean(name = 'thermal',limits=c(10,25),na.value = NA) +
  ggtitle(paste("Satellite SST and buoy temperature (circles) \n", unique(as.Date(SST$time)))) +
  geom_point(data=buoy2, aes(x=lon,y=lat,color=temp.day),size=3,shape=21,color="black") + 
  scale_color_cmocean(name = 'thermal',limits=c(10,25),na.value ="grey20") 
  
myplot
```

## References:
- <a href="https://coastwatch.pfeg.noaa.gov/data.html">NOAA CoastWatch Westcoast Node Data Catalog</a>
- <a href="https://www.ndbc.noaa.gov/download_data.php?filename=46259h2017.txt.gz&dir=data/historical/stdme ">NOAA National Data Buoy Center</a>
- <a href="https://docs.ropensci.org/rerddap/">Rerddap R Package reference</a>

