---
title: "subset-polar-data-with-shapefile"
author: "Sunny Hospital"
date: "`r Sys.Date()`"
output: html_document
---

# https://rspatial.org/spatial/index.html

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
# Function to check if pkgs are installed, install missing pkgs, and load

pkgTest <- function(x)
{
  if (!require(x,character.only = TRUE))
  {
    install.packages(x,dep=TRUE,repos='http://cran.us.r-project.org')
    if(!require(x,character.only = TRUE)) stop(x, " :Package not found")
  }
}

list.of.packages <- c("sp", "ggplot2" , "rerddap", "RColorBrewer", "scales", "reshape2")

# create list of installed packages
pkges = installed.packages()[,"Package"]
for (pk in list.of.packages) {
  pkgTest(pk)
}
```


### The exercise demonstrates the following techniques:
- Loading twice daily sea ice thickness data of the year 2023 from ERDDAP using `xarray` and compute monthly average
- Loading multi-year monthly average sea ice thickness data from netCDF file
- Calculating the trend of monthly sea ice thickness means for the climatological reference period of 2006 to 2020 (15 years)
- Calculating historical average of the reference period (2006-2020)
- Calculating anomalies (departures from the historical mean)
- Visualizing the data 


```{r}
library(terra)
library(sf)
library(ggplot2)
library(rerddap)
```


## Load IMS Sea ice data 

```{r erddap_data}

pw_url = "https://polarwatch.noaa.gov/erddap/"
dataset_id = "usnic_ims_4km"
var_name = "IMS_Surface_Values"
dat_info <- info(datasetid = dataset_id, url = pw_url)
dat_info
```

```{r}
dat <- griddap(dat_info,  time = c('2019-11-01','2019-11-01'), fields = var_name)
ims <- dat$data

```


# Visualize the IMS satellite data

```{r, eval=FALSE }
# Visualize the data (warning: this takes awhile to )
ggplot(data = ims, aes(x = x, y = y, fill=IMS_Surface_Values)) +
         geom_tile() + 
         coord_fixed(ratio = 1) + 
         scale_y_continuous() + 
         scale_x_continuous() +
         scale_fill_gradientn(colours=rev(brewer.pal(n = 3, name = "Blues")),na.value="tan") +
         ggtitle("IMS Data")

```

## Convert IMS data to raster S4 object

```{r}

# convert ims to raster S3 obj
ims_ras <- terra::rast(ims)

# plot the raster data
plot(ims_ras)

# get CRS of the ims data
data_crs <- crs(ims_ras)
```
# Load the lake shapefile

```{r read_shapefile}
# Set directory path for shapefile
dir_path <- '../resources/Iliamna/Iliamna.shp'


# Load the shape file
shapes <- st_read(dsn = dir_path)

# View
print(shapes)

# Get boundary coordinates for Lake Iliamna
lake_shp <- shapes[shapes$Lake_name == "Iliamna", ]
ggplot(data = lake_shp) +
  geom_sf() +
  theme_minimal() +
  labs(title = "Lake Iliamna", 
       caption = "Source: shapefile")

```
## Convert the lake shape to raster S4 object

```{r}

shapes_polar <- st_transform(lake_shp, data_crs)
lake_ras <- terra::vect(shapes_polar)
plot(lake_ras)
```


## Crop the IMS data

```{r}

# crop to the bounding box of the lake
cropped_lake <- crop(ims_ras, lake_ras)

# mask data with the lake shape
masked_lake <- mask(cropped_lake, lake_ras)
plot(masked_lake)

```
