---
title: "Calculate sea ice extent"
author: "NOAA CoastWatch"
date updated: "8/30/2023"
output:
  md_document:
    variant: gfm
---

# Define a marine habitat
>notebook filename | define_marine_habitat.Rmd  

The TurleWatch project investigated the thermal habitat of loggerhead sea turtles in the Pacific Ocean north of the Hawaiian Islands. Research results indicate that most loggerhead turtles stay in water between 17.5째C and 18.5째C. When the 17.5째C to 18.5째C temperature contour is drawn on a map of sea surface temperature conditions, it delineates the boundary of the loggerhead's preferred habitat.  

In this exercise you will plot the thermal boundary of loggerhead sea turtles using satellite sea surface temperature. The exercise demonstrates the following techniques:  
* Using xtracto_3D to extract data from a rectangular area  
* Masking a data array  
* Plotting maps using ggplot  

## Install required packages and load libraries 

```{r install, message=FALSE, warning=FALSE}

# Function to check if pkgs are installed, install missing pkgs, and load
pkgTest <- function(x)
{
  if (!require(x,character.only = TRUE))
  {
    install.packages(x,dep=TRUE,repos='http://cran.us.r-project.org')
    if(!require(x,character.only = TRUE)) stop(x, " :Package not found")
  }
}

list.of.packages <- c( "ncdf4", "rerddap", "plotdap", "RCurl",  
                       "raster", "colorRamps", "maps", "mapdata",
                       "ggplot2", "RColorBrewer", "rerddapXtracto")

# create list of installed packages
pkges = installed.packages()[,"Package"]

for (pk in list.of.packages) {
  pkgTest(pk)
}


```

## Select the Satellite Data
* Use the MUR SST dataset (ID jplMURSST41mday)  
* Gather information about the dataset (metadata) using **rerddap**  
* Displays the information  
```{r dataInfo}
# CHOOSE DATASET and get information about it 


url = "http://coastwatch.pfeg.noaa.gov/erddap/"
dataInfo <- rerddap::info('jplMURSST41mday',url=url)
parameter <- 'sst'
```

## Get Satellite Data  
* Select an area off the coast of California: longitude range of -130 to -115 east  and latitude range of 25 to 40 north  
* Set the time range to days withing one month: tcoord=c('2018-06-06','2018-06-08')). The values do have to be different.  

```{r getdata}

# latitude and longitude of the vertices
ylim<-c(25,40)
xlim<-c(-130,-115)

# Choose an area off the coast of California
# Extract the data
SST <- rxtracto_3D(dataInfo,xcoord=xlim,ycoord=ylim,parameter=parameter, 
                   tcoord=c('2018-06-06','2018-06-08'))

# Drop command needed to reduce SST from a 3D variable to a 2D  one  
SST$sst <- drop(SST$sst) 
```
## Make a quick plot using plotBBox  
```{r qplot}

plotBBox(SST, plotColor = 'thermal',maxpixels=100000)

```

## Define the Thermal niche of Loggerhead Turtles  

__ Set the thermal range to 17.5-18.5 degrees C, as determined by the TurtleWatch program.__  

```{r temp}
## Define turtle temperature range
min.temp <- 17.5
max.temp <- 18.5
```
__Create another variable for habitat temperature__

Set the habitat temperature to equal NA 

```{r makeVar}

SST2 <- SST
SST2$sst[SST2$sst >= min.temp & SST2$sst <= max.temp] <- NA
plotBBox(SST2, plotColor = 'thermal',maxpixels=100000)

```

It would be nicer to color in the turtle habitat area (the NA values) with a different color. If you want to customize graphs its better to use `ggplot` than the `plotBBox` that comes with 
`rerrdapXtracto` package.  Here we will use `ggplot` to plot the data. But first the data is reformatted for use in `ggplot`. 

Restructure the data 
```{r restructure}

dims <- dim(SST2$sst)
SST2.lf <- expand.grid(x=SST$longitude,y=SST$latitude)
SST2.lf$sst<-array(SST2$sst,dims[1]*dims[2])

```

## Plot the Data using 'ggplot'  
```{r plot}
coast <- map_data("worldHires", ylim = ylim, xlim = xlim)

par(mar=c(3,3,.5,.5), las=1, font.axis=10)

myplot<-ggplot(data = SST2.lf, aes(x = x, y = y, fill = sst)) +
  geom_tile(na.rm=T) +
  geom_polygon(data = coast, aes(x=long, y = lat, group = group), fill = "grey80") +
  theme_bw(base_size = 15) + ylab("Latitude") + xlab("Longitude") +
  coord_fixed(1.3,xlim = xlim, ylim = ylim) +
  ggtitle(unique(as.Date(SST2$time))) +
  scale_fill_gradientn(colours = rev(rainbow(12)),limits=c(10,22),na.value = "firebrick4") 

myplot

```



